---
title: "ETC5521: Exploratory Data Analysis"
subtitle: "Working with a single variable, making transformations, detecting outliers, using robust statistics"
author: "Emi Tanaka"
email: "ETC5521.Clayton-x@monash.edu"
date: "Week 4 - Session 2"
length: "50 minutes"
titlebgimg: "images/bg-01.png"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/font-awesome-all.css"
      - "assets/tachyons-addon.css"
      - "assets/animate.css"
      - "assets/fira-code.css"
      - "assets/boxes.css"
      - "assets/table.css"
      - "assets/styles.css"
      - "assets/monash-brand.css"
      - "assets/monash-fonts.css"
      - "assets/slide-types.css"
      - "assets/panelset-modified.css"
      - "assets/scroll.css"
    self_contained: false 
    seal: false 
    chakra: 'lib/remark-latest.min.js'
    lib_dir: lib
    includes:
      in_header: "assets/head.html"
    mathjax: "lib/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: magula
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---

```{r, include = FALSE}
current_file <- knitr::current_input()
basename <- gsub(".Rmd$", "", current_file)
```

```{r, include = FALSE}
library(tidyverse)
library(colorspace)
library(patchwork)
options(width = 200)
knitr::opts_chunk$set(
  fig.path = "images/week4B/",
  fig.width = 6,
  fig.height = 3.5,
  fig.align = "center",
  dev.args = list(bg = 'transparent'),
  #out.width = "100%",
  fig.retina = 3,
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = TRUE,
  cache.path = "cache/week4B/"
)
theme_set(ggthemes::theme_gdocs(base_size = 12) +
            theme(plot.background = element_rect(fill = 'transparent', colour = NA), axis.line.y = element_line(color = "black", linetype = "solid"),
                  plot.title.position = "plot",
                  plot.title = element_text(size = 16),
                  panel.background  = element_rect(fill = 'transparent', colour = NA),
                  legend.background = element_rect(fill = 'transparent', colour = NA),
                  legend.key        = element_rect(fill = 'transparent', colour = NA)
                  ) )
```


```{r titleslide, child="assets/titleslide.Rmd"}
```


---

class: transition 

# Bins and Bandwidths


---


# .orange[Case study] .circle.bg-orange.white[3] Boston housing data .f4[Part 1/4]


```{r boston-data, include = FALSE}
data(bostonc, package = "DAAG")
df3 <- read_tsv(bostonc[10:length(bostonc)]) 
skimr::skim(df3)
```

```{r boston-data-longer, include = FALSE}
df3long <- df3 %>% pivot_longer(MEDV:LSTAT,
                             names_to = "var",
                             values_to = "value") %>% 
  filter(!var %in% c("CHAS", "B", "ZN"))
skimr::skim(df3long)
```


.panelset[
.panel[.panel-name[📊]
.grid[
.item[
```{r boston-plot1, fig.width = 6.4}
ggplot(df3, aes(MEDV)) + 
  geom_histogram(binwidth = 1, color = "black", fill = "#008A25") + 
  labs(x = "Median housing value (US$1000)", y = "Frequency")

```

]
.item[

{{content}}


]
]
]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r boston-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f5[
```{r, ref.label = c("boston-plot1"), echo = TRUE, eval = FALSE}
```

]]
]

.footnote.f6[
Harrison, David, and Daniel L. Rubinfeld (1978) Hedonic Housing Prices and the Demand for Clean Air, *Journal of Environmental Economics and Management* **5** 81-102. Original data.<br>
Gilley, O.W. and R. Kelley Pace (1996) On the Harrison and Rubinfeld Data. *Journal of Environmental Economics and Management* **31** 403-405. Provided corrections and examined censoring.<br>
Maindonald, John H. and Braun, W. John (2020). DAAG: Data Analysis and Graphics Data and Functions. R package version 1.24
]

--

* Thre is a large frequency in the final bin.
* There is a decline in observations in the $40-49K range as well as dip in observations around $26K and $34K.
* The histogram is using a bin width of 1 unit and is **left-open** (or **right-closed**): (4.5, 5.5], (5.5, 6.5] ... (49.5, 50.5].
* Occasionally, whether it is left- or right-open can make a difference.


---

# .orange[Case study] .circle.bg-orange.white[3] Boston housing data .f4[Part 2/4]



.panelset[
.panel[.panel-name[📊]
.grid[
.item[
```{r boston-plot2, fig.height = 2}
ggplot(df3, aes(MEDV, y = "")) + 
  geom_boxplot(fill = "#008A25") + 
  labs(x = "Median housing value (US$1000)", y = "") + 
  theme(axis.line.y = element_blank())

```
```{r boston-plot3, fig.height = 2}
ggplot(df3, aes(MEDV, y = "")) + 
  geom_jitter() + 
  labs(x = "Median housing value (US$1000)", y = "") + 
  theme(axis.line.y = element_blank())

```
```{r boston-plot4, fig.height = 3.5}
ggplot(df3, aes(MEDV)) + 
  geom_density() + 
  geom_rug() + 
  labs(x = "Median housing value (US$1000)", y = "") + 
  theme(axis.line.y = element_blank())

```
]
.item[

* Density plots depend on the bandwidth chosen and more than often do not estimate well at boundary cases
* There are various way to present features of the data using a plot and what works for one person, may not be as straightforward for another
* Be prepared to do multiple plots!

]
]
]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r boston-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r, ref.label = paste0("boston-plot", 2:4), echo = TRUE, eval = FALSE}
```
]
]
]

---

# .orange[Case study] .circle.bg-orange.white[3] Boston housing data .f4[Part 3/4]



.panelset[
.panel[.panel-name[📊]
.grid[
.item[
```{r boston-plot5, fig.height = 2}
ggplot(df3, aes(PTRATIO)) + 
  geom_histogram(fill = "#9651A0",  color = "black", binwidth = 0.2) + 
  labs(x = "Pupil-teacher ratio by town", y = "",
       title = "Bin width = 0.2, Left-open") 

```
```{r boston-plot6, fig.height = 2}
ggplot(df3, aes(PTRATIO)) + 
  geom_histogram(fill = "#9651A0",  color = "black", binwidth = 0.5) + 
  labs(x = "Pupil-teacher ratio by town", y = "",
       title = "Bin width = 0.5, Left-open") 

```
```{r boston-plot7, fig.height = 2}
ggplot(df3, aes(PTRATIO)) + 
  geom_histogram(fill = "#9651A0",  color = "black", bins = 30) + 
  labs(x = "Pupil-teacher ratio by town", y = "",
       title = "Bin number = 30, Left-open") 

```


]
.item[

```{r boston-plot8, fig.height = 2}
ggplot(df3, aes(PTRATIO)) + 
  geom_histogram(fill = "#9651A0", color = "black", binwidth = 0.2, closed = "left") + 
  labs(x = "Pupil-teacher ratio by town", y = "",
       title = "Bin width = 0.2, Right-open") 

```
```{r boston-plot9, fig.height = 2}
ggplot(df3, aes(PTRATIO)) + 
  geom_histogram(fill = "#9651A0", color = "black", binwidth = 0.5, closed = "left") + 
  labs(x = "Pupil-teacher ratio by town", y = "",
       title = "Bin width = 0.5, Right-open") 

```
```{r boston-plot10, fig.height = 2}
ggplot(df3, aes(PTRATIO)) + 
  geom_histogram(fill = "#9651A0", color = "black",
                 bins = 30, closed = "left") + 
  labs(x = "Pupil-teacher ratio by town", y = "",
       title = "Bin number = 30, Right-open") 

```



]
]
]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r boston-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4.scroll-sign[.s500[
```{r, ref.label = paste0("boston-plot", 5:10), echo = TRUE, eval = FALSE}
```
]]]
]

---

# .orange[Case study] .circle.bg-orange.white[3] Boston housing data .f4[Part 4/4]


.panelset[
.panel[.panel-name[📊]
.grid[
.item[
```{r boston-plotx, fig.height = 6, fig.width = 8}
ggplot(df3long, aes(value)) +
  geom_histogram(color = "white") +
  facet_wrap( ~var, scale = "free") + 
  labs(x = "", y = "") + 
  theme(axis.text = element_text(size = 12))

```

]
.item.f4[

* CRIM: per capita crime rate by town 
* INDUS: proportion of non-retail business acres per town
* NOX: nitrogen oxides concentration (parts per 10 million)
* RM: average number of room per dwelling
* AGE: proportion of owner-occupied units built prior to 1940
* DIS: weighted mean of distances to 5 Boston employment centres
* RAD: index of accessibility to radial highways
* TAX: full-value property tax rate per $10K
* PTRATIO: pupil-teacher ratio by town 
* LSTAT: lower status of the population (%)
* MEDV: median value of owner-occupied homes in $1000s

]
]
]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r boston-data-longer, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r, ref.label = c("boston-plotx"), echo = TRUE, eval = FALSE}
```
]
]
]


---


# .orange[Case study] .circle.bg-orange.white[4] Hidalgo stamps thickness 

```{r hidalgo-data, include = FALSE}
load(here::here("data/Hidalgo1872.rda"))
skimr::skim(Hidalgo1872)
```


.panelset[
.panel[.panel-name[📊]
```{r hidalgo-plot, fig.height = 4, fig.width = 8}
ggplot(Hidalgo1872, aes(thickness)) +
  geom_histogram(binwidth = 0.001, aes(y = stat(density))) + 
  labs(x = "Thickness (0.001 mm)", y = "Density") + 
  geom_density(color = "#E16A86", size = 2) + 
  geom_density(color = "#00AD9A", size = 2, bw = "SJ")

```

* A stamp collector, Walton von Winkle, bought several collections of Mexican stamps from 1872-1874 and measured the thickness of all of them.
* The different bandwidth for the density plot suggest either that there are two or seven modes.

]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r hidalgo-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r, ref.label = c("hidalgo-plot"), echo = TRUE, eval = FALSE}
```
]
]
]


---

class: transition 


# Focus

---

# .orange[Case study] .circle.bg-orange.white[5] Movie length

```{r movies-data, include = FALSE}
data(movies, package = "ggplot2movies")
skimr::skim(movies)
```


.panelset[
.panel[.panel-name[📊]
.grid[
.item[
```{r movies-plot1}
ggplot(movies, aes(length)) +
  geom_histogram(color = "white") + 
  labs(x = "Length of movie (minutes)", y = "Frequency")

```
```{r movies-plot2, fig.width = 5.3}
ggplot(movies, aes(length)) +
  geom_histogram(color = "white") + 
  labs(x = "Length of movie (minutes)", y = "Frequency") + 
  scale_x_log10()

```

]
.item[

{{content}}
]
]


]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r movies-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r, ref.label = paste0("movies-plot", 1:3), echo = TRUE, eval = FALSE}
```
]
]
]

--

* Upon further exploration, you can find the two movies that are well over 16 hours long are `r knitr::combine_words(paste0('"<i>', str_replace(pull(filter(movies, length > 1000), title), ", The", ""), '</i>"'))`

{{content}}
--

* We can restrict our attention to films under 3 hours:

```{r movies-plot3, fig.width = 9}
movies %>% 
  filter(length < 180) %>% 
  ggplot(aes(length)) +
  geom_histogram(binwidth = 1, fill = "#795549", color = "black") + 
  labs(x = "Length of movie (minutes)", y = "Frequency") 

```

{{content}}

--

* Notice that there is a peak at particular times. Why do you think so?


---



class: transition middle

# Categorical variables

<br><br>


This lecture is based on Chapter 4  of <br><br>Unwin (2015) Graphical Data Analysis with R

---

class: middle

# There are two types of categorical variables

--

<br><br>

.monash-blue[**Nominal**] where there is no intrinsic ordering to the categories<br>
**E.g.** blue, grey, black, white.

--

<br>

.monash-blue[**Ordinal**] where there is a clear order to the categories.<Br>
**E.g.** Strongly disagree, disagree, neutral, agree, strongly agree.


---

# Categorical variables in R

.grid[
.item.br[
* In R, categorical variables may be encoded as **factors**.
.f4[
```{r factors, echo = TRUE}
data <- c(2, 2, 1, 1, 3, 3, 3, 1)
factor(data)
```
]
* You can easily change the labels of the variables:
.f4[
```{r factor-labels, echo = TRUE}
factor(data, labels = c("I", "II", "III"))
```
]
]
.item.f4[

{{content}}

]
]

--

* Order of the factors are determined by the input:


```{r factor-input, echo = TRUE}
# numerical input are ordered in increasing order #<<
factor(c(1, 3, 10))
# character input are ordered alphabetically #<<
factor(c("1", "3", "10"))
# you can specify order of levels explicitly #<<
factor(c("1", "3", "10"),  
       levels = c("1", "3", "10"))
```


---

# Numerical factors in R

```{r, echo = TRUE, warning = TRUE}
x <- factor(c(10, 20, 30, 10, 20))
mean(x)
```

--

<i class="fas fa-exclamation-triangle"></i> `as.numeric` function returns the internal integer values of the factor
```{r, echo = TRUE}
mean(as.numeric(x))
```

--

You probably want to use:

.flex[
.w-50[
```{r, echo = TRUE}
mean(as.numeric(levels(x)[x]))
```

]
.w-50[
```{r, echo = TRUE}
mean(as.numeric(as.character(x)))
```

]
]

---

# .orange[Revisiting Case study] .circle.bg-orange.white[1] 2019 Australian Federal Election

```{r aus-election-data1}
df1 <- read_csv(here::here("data/HouseFirstPrefsByCandidateByVoteTypeDownload-24310.csv"), 
                skip = 1,
                col_types = cols(
                      .default = col_character(),
                      OrdinaryVotes = col_double(),
                      AbsentVotes = col_double(),
                      ProvisionalVotes = col_double(),
                      PrePollVotes = col_double(),
                      PostalVotes = col_double(),
                      TotalVotes = col_double(),
                      Swing = col_double()))
```
```{r aus-election-data3}
tdf3 <- df1 %>% 
  group_by(DivisionID) %>% 
  summarise(DivisionNm = unique(DivisionNm),
            State = unique(StateAb),
            votes_GRN = TotalVotes[which(PartyAb=="GRN")],
            votes_total = sum(TotalVotes)) %>% 
  mutate(perc_GRN = votes_GRN / votes_total * 100)
```
.panelset[
.panel[.panel-name[📊]

.flex[
.w-50[
```{r aus-election-plot1, fig.height = 5}
tdf3 %>% 
  ggplot(aes(perc_GRN, State)) +
  ggbeeswarm::geom_quasirandom(groupOnX = FALSE, varwidth = TRUE) +
  labs(x = "Percentage of first preference votes per division",
       y = "State", 
       title = "First preference votes for the Greens party")
```




]
.w-50[
```{r aus-election-plot2, fig.height = 5}
tdf3 %>% 
  mutate(State = fct_reorder(State, perc_GRN)) %>% 
  ggplot(aes(perc_GRN, State)) +
  ggbeeswarm::geom_quasirandom(groupOnX = FALSE, varwidth = TRUE) +
  labs(x = "Percentage of first preference votes per division",
       y = "State", 
       title = "First preference votes for the Greens party")
```
]

]

]
.panel[.panel-name[data]
.scroll-sign[.s400.f4[
```{r, ref.label = c("aus-election-data1", "aus-election-data3"), echo = TRUE}
```
]]]
.panel[.panel-name[R]
.f4[
```{r, ref.label=paste0("aus-election-plot", 1:2), echo = TRUE, eval = FALSE}
```
]]
]


---

class:  middle

# Order nominal variables meaningfully

<i class="fas fa-code"></i> **Coding tip**: use below functions to easily change the order of factor levels

```{r, eval = FALSE, echo = TRUE}
stats::reorder(factor, value, mean)
forcats::fct_reorder(factor, value, median)
forcats::fct_reorder2(factor, value1, value2, func)
```






---

# .orange[Case study] .circle.bg-orange.white[6] Aspirin use after heart attack

```{r meta-data, include = FALSE}
data("Fleiss93", package = "meta")
df6 <- Fleiss93 %>% 
  mutate(total = n.e + n.c)
skimr::skim(df6)
```

.panelset[
.panel[.panel-name[📊]

.grid[
.item[
```{r meta-plot1, fig.height = 3}
df6 %>% 
  mutate(study = fct_reorder(study, desc(total))) %>% 
  ggplot(aes(study, total)) + 
  geom_col() + 
  labs(x = "", y = "Frequency") + 
  guides(x = guide_axis(n.dodge = 2))

```

```{r meta-plot2, fig.height = 3}
df6 %>% 
  mutate(study = ifelse(total < 2000, "Other", study),
         study = fct_reorder(study, desc(total))) %>% 
  ggplot(aes(study, total)) + 
  geom_col() + 
  labs(x = "", y = "Frequency") 

```

]
.item[

* Meta-analysis is a statistical analysis that combines the results of multiple scientific studies.
* This data studies the use of aspirin for death prevention after myocardial infarction, or in plain terms, a heart attack.
* The ISIS-2 study has more patients than all other studies combined.
* You could consider lumping the categories with low frequencies together.
]


]

]
.panel[.panel-name[data]
.h300.f4.scroll-sign[
```{r meta-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r , ref.label=paste0("meta-plot", 1:2), echo = TRUE, eval = FALSE}
```
]]
]

.f5.footnote[
Fleiss JL (1993): The statistical basis of meta-analysis. *Statistical Methods in Medical Research* **2** 121–145<br>
Balduzzi S, Rücker G, Schwarzer G (2019), How to perform a meta-analysis with R: a practical tutorial, Evidence-Based Mental Health.
]

---

class: nostripheader middle


# Consider combining factor levels with low frequencies

<i class="fas fa-code"></i> **Coding tip**: the following family of functions help to easily lump factor levels together:

```{r fct_lump, echo = TRUE, eval = FALSE}
forcats::fct_lump()
forcats::fct_lump_lowfreq()
forcats::fct_lump_min()
forcats::fct_lump_n()
forcats::fct_lump_prop()
# if conditioned on another variable
ifelse(cond, "Other", factor)
dplyr::case_when(cond1 ~ "level1",
                 cond2 ~ "level2",
                 TRUE ~ "Other")
```


---

# .orange[Case study] .circle.bg-orange.white[7] Anorexia

```{r anorexia-data, include = FALSE}
data(anorexia, package = "MASS")

df9tab <- table(anorexia$Treat) %>% 
  as.data.frame() %>% 
  rename(Treatment = Var1, Frequency = Freq)

skimr::skim(anorexia)
```

.panelset[
.panel[.panel-name[📊]

.grid[
.item[
```{r anorexia-plot1}
ggplot(anorexia, aes(Treat)) + 
  geom_bar() + 
  labs(x = "", y = "Frequency") 

```

```{r anorexia-table}
df9tab %>% 
  knitr::kable() %>% 
  kableExtra::kable_classic(full_width = FALSE)
```

]
.item[

**Table or Plot?**

{{content}}

]


]

]
.panel[.panel-name[data]
.h200.f4.scroll-sign[
```{r anorexia-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r , ref.label="anorexia-plot1", echo = TRUE, eval = FALSE}
```
```{r anorexia-plot2, echo = TRUE, eval = FALSE}
ggplot(anorexia, aes(Treat)) + 
  stat_count(geom = "point", size = 4) +
  stat_count(geom = "line", group = 1) +
  labs(y = "Frequency", x = "")
```

]]
]

.f6.footnote[

Hand, D. J., Daly, F., McConway, K., Lunn, D. and Ostrowski, E. eds (1993) A Handbook of Small Data Sets. Chapman & Hall, Data set 285 (p. 229) <br>
Venables, W. N. & Ripley, B. D. (2002) Modern Applied Statistics with S. Fourth Edition. Springer, New York. ISBN 0-387-95457-0

]

--

* Table for accuracy, plot for visual communication

{{content}}

--

**Why not a point or line?**

```{r anorexia-plot2, fig.height = 2.5}
```

{{content}}

--

<div class="f4">
<ul>
<li>This can be appropriate depending on what you want to communicate </li>
<li>A barplot occupies more area compared to a point and the area does a better job of communicating size</li>
<li>A line is suggestive of a trend </li>
</ul>
</div>

---


# .orange[Case study] .circle.bg-orange.white[8] Titanic

```{r titanic-data, include = FALSE}
df9 <- as_tibble(Titanic)
skimr::skim(df9)
```

.panelset[
.panel[.panel-name[📊]

.flex[
.w-50[
.flex[
.w-50[
```{r titanic-plot1, fig.height = 2, fig.width = 4}
df9 %>% 
  group_by(Class) %>% 
  summarise(total = sum(n)) %>% 
  ggplot(aes(Class, total)) + 
  geom_col(fill = "#ee64a4") + 
  labs(x = "", y = "Frequency") 

```

```{r titanic-plot2, fig.height = 2, fig.width = 3}
df9 %>% 
  group_by(Sex) %>% 
  summarise(total = sum(n)) %>% 
  ggplot(aes(Sex, total)) + 
  geom_col(fill = "#746FB2") + 
  labs(x = "", y = "Frequency") 

```

]
.w-50[
```{r titanic-plot3, fig.height = 2, fig.width = 3}
df9 %>% 
  group_by(Age) %>% 
  summarise(total = sum(n)) %>% 
  ggplot(aes(Age, total)) + 
  geom_col(fill = "#C8008F") + 
  labs(x = "", y = "Frequency") 

```

```{r titanic-plot4, fig.height = 2, fig.width = 3}
df9 %>% 
  group_by(Survived) %>% 
  summarise(total = sum(n)) %>% 
  ggplot(aes(Survived, total)) + 
  geom_col(fill = "#795549") + 
  labs(x = "Survived", y = "Frequency") 

```

]
]


]
.w-50[

**What does the graphs for each categorical variable tell us?**

* There were more crews than 1st to 3rd class passengers
* There were far more males on ship; possibly because majority of crew members were male. You can further explore this by constructing two-way tables or graphs that consider both variables.
* Most passengers were adults. 
* More than two-thirds of passengers died. 

]


]

]
.panel[.panel-name[data]
.h350.f4.scroll-sign[
```{r titanic-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4.scroll-sign[.s500[
```{r , ref.label=paste0("titanic-plot", 1:4), echo = TRUE, eval = FALSE}
```
]]
]]

.f5.footnote[
British Board of Trade (1990), Report on the Loss of the ‘Titanic’ (S.S.). British Board of Trade Inquiry Report (reprint). Gloucester, UK: Allan Sutton Publishing
]


---


class: nostripheader middle

# Coloring bars


```{r, fig.width = 10}
g1 <- df9 %>% 
  group_by(Class) %>% 
  summarise(total = sum(n)) %>% 
  ggplot(aes(Class, total)) + 
  geom_col() + 
  labs(x = "", y = "Frequency") 

g2 <- df9 %>% 
  group_by(Class) %>% 
  summarise(total = sum(n)) %>% 
  ggplot(aes(Class, total)) + 
  geom_col(fill = "#006DAE") + 
  guides(fill = FALSE) + 
  labs(x = "", y = "Frequency") + 
  scale_fill_discrete_qualitative()

g3 <- df9 %>% 
  group_by(Class) %>% 
  summarise(total = sum(n)) %>% 
  ggplot(aes(Class, total)) + 
  geom_col(aes(fill = Class)) + 
  guides(fill = FALSE) + 
  labs(x = "", y = "Frequency") + 
  scale_fill_discrete_qualitative()

g1 + g2 + g3
```

--

* Colour here doesn't add information as the x-axis already tells us about the categories, but colouring bars can make it more visually appealing. 
* If you have too many categories colour won't work well to differentiate the categories. 

---

# .orange[Case study] .circle.bg-orange.white[9] Opinion poll in Ireland Aug 2013

```{r poll-data, include = FALSE}
df9 <- tibble(party = c("Fine Gael", "Labour", "Fianna Fail",
                         "Sinn Fein", "Indeps", "Green", "Undecided"),
               nos = c(181, 51, 171, 119, 91, 4, 368)) 
df9v2 <- df9 %>% filter(party != "Undecided")
df9
```

.panelset[
.panel[.panel-name[📊]

.grid[
.item[
```{r poll-plot1, fig.width = 7}
g9 <- df9 %>% 
  ggplot(aes("", nos, fill = party)) + 
  geom_col(color = "black") + 
  labs(y = "", x = "") + 
  coord_polar("y") +
  theme(axis.line = element_blank(),
        axis.line.y = element_blank(),
        axis.text = element_blank(),
        panel.grid.major = element_blank()) +
  scale_fill_discrete_qualitative(name = "Party")
g9

```

```{r poll-plot2, fig.width = 7}
g9 %+% df9v2 + 
  # below is needed to keep the same color scheme as before
  scale_fill_manual(values = qualitative_hcl(7)[1:6])

```

]
.item[

* Pie chart is popular in mainstream media but are not generally recommended as people are generally poor at comparing angles.
* 3D pie charts should definitely be avoided!
* Here you can see that there are many people that are "Undecided" for which political party to support and failing to account for this paints a different picture. 

]


]

]
.panel[.panel-name[data]
.f4[
```{r poll-data, echo = TRUE, render = knitr::normal_print}
```
]]
.panel[.panel-name[R]
.f4[
```{r , ref.label=paste0("poll-plot", 1:2), echo = TRUE, eval = FALSE}
```
]]]




---

class: middle

# Piechart is a stacked barplot just with a transformed coordinate system

--

```{r barplot, echo = TRUE, fig.height = 2, fig.width = 3}
df <- data.frame(var = c("A", "B", "C"), perc = c(40, 40, 20))
g <- ggplot(df, aes("", perc, fill = var)) + 
  geom_col()
g
```
--

```{r piechart, echo = TRUE, fig.height = 2, fig.width = 3}
g + coord_polar("y")
```

---

class: middle

# Roseplot is a barplot just with a transformed coordinate system

--

```{r nonstacked-barplot, echo = TRUE, fig.height = 2, fig.width = 10}
dummy <- data.frame(var = LETTERS[1:20], 
                 n = round(rexp(20, 1/100)))
g <- ggplot(dummy, aes(var, n)) + geom_col(fill = "pink", color = "black")
g
```
--

```{r roseplot, echo = TRUE, fig.height = 2, fig.width = 3}
g + coord_polar("x") + theme_void()
```

---

# Take away messages

.flex[
.w-70.f2[



<ul class="fa-ul">
{{content}}
</ul>


]
]

--

<li><span class="fa-li"><i class="fas fa-paper-plane"></i></span>Again, be prepared to do multiple plots</li>
{{content}}
--

<li><span class="fa-li"><i class="fas fa-paper-plane"></i></span>Changing bins or bandwidth in histogram, violin or density plots can paint a different picture</li>
{{content}}
--

<li><span class="fa-li"><i class="fas fa-paper-plane"></i></span>Consider different representations of categorical variables (reordering meaningfully, lumping low frequencies together, plot or table, pie or barplot, missing categories)</li>

---

```{r endslide, child="assets/endslide.Rmd"}
```
